Taking into consideration both the previous work in the field and the experience of our laboratory in degradomics, in this thesis we aimed to annotate, analyse and compare proteases and other genes in the genomes of long-lived metazoans.

Thus, the specific objectives for this Doctoral Thesis were:

\begin{enumerate}[topsep=1ex,itemsep=1ex]
    \item Study and analysis of \textsl{Physeter macrocephalus}'s Degradome.
    \begin{enumerate}[label=\arabic{enumi}.\arabic*,itemsep=0ex]
        \item Manual annotation of \textsl{P. macrocephalus}'s Degradome.
        \item Comparison of this gene set with that of the other annotated marine mammals (i.e. Bowhead whale, Minke whale, Bottlenose dolphin, and Killer whale).
        \item Study of the differences and similarities in an evolutionary context.
    \end{enumerate}
    \item Study and analysis of \textsl{C. abingdonii}'s Degradome.
    \begin{enumerate}[label=\arabic{enumi}.\arabic*,itemsep=0ex]
        \item Manual annotation of \textsl{C. abingdonii}'s Degradome.
        \item Comparison of this gene set with that of the other annotated tortoises (i.e. other Giant Galapagos tortoises, continental relatives, and Aldabra tortoise).
        \item Study of the differences and similarities in an evolutionary context.
    \end{enumerate}
    \item Study and analysis of other genomic families of interest in \textsl{C. abingdonii}.
    \begin{enumerate}[label=\arabic{enumi}.\arabic*,itemsep=0ex]
        \item Automatic annotation of the complete genome of \textsl{C. abingdonii}.
        \item Study of the evolutionary traits of the different families.
        \item Study of the genomic variation of gene families of interest among related species.
    \end{enumerate}
\end{enumerate}
